# Introduction to Demographic Heterogeneity {#intro}

## Outline
1. What demographic heterogeneity is (and isn't)
2. Dynamics of population growth with two sub-groups
3. Keyfitz's result $\bar{r}'(t) = \sigma^2_r(t)$.
4. Ken's model of Poisson heterogeneity

Resources:

Content is based on Section 8.8 from @wachter2014essential as well as @vaupel1985heterogeneity for an accessible introduction to heterogeneity and overview. Derivations also from @keyfitz2005applied.

## Part I. Conceptual Introduction

### What is Demographic Heterogeneity?
  
- If we see different outcomes (e.g., people dying at different ages), is this *Demographic Heterogeneity*? NO.
- Demographic heterogeneity $=$ different rates for different folks.

>In a *demographically heterogeneous* population, people are of different types, with different type-specific rates.

These types can be discrete, with individuals being homogeneous within their type, or they can be continuous with possibly no individual having exactly the same risk as another.
  
*NOTE: maybe add here a summary of @vaupel1985heterogeneity?*

### An example
  
  Let's draw 10 individuals from a homogeneous population and heterogeneous population.

```{r hetero-homo, fig.cap='Homogeneity and Heterogeneity', out.width='80%', fig.align='center', fig.height=4}
  ## Homogeneous hazard of 1/10
  set.seed(13)
  x.homo <- rexp(10, rate = 1/10)
  ## Heterogeneous hazard (half 1/6 and half 1/13)
  ## Note: I didn't pick these particular numbers for any specific reason
  x.hetero <- c(rexp(5, rate = 1/6),
                rexp(5, rate = 1/13))
  par(mfrow = c(1,2))
  dotchart(x.homo, main = "homogenous variation")
  dotchart(x.hetero, main = "heterogeneous variation")
```

  
- Can you tell which is which?
  - _Homogeneous_: Chance only
  - *Heterogeneous:* Chance + group variation in risk
- Would we expect to see a difference if we increased sample?

ANSWER THIS
  
- Analogies ?
  - Social inequality: equal opportunity vs. equal outcomes
  - Analysis of variance $\mbox{total variance} = \mbox{within group} + \mbox{between group}$
  - Statistical models $y = a + b x + \epsilon$
  

- What's new? Dynamics.

  Heterogeneous populations evolve differently. Aggregates $\neq$ Individuals
  
  - Rates of growth (or decline)
  - Changes over time or age or duration
  - The trajectory of even the \alert{average} individual differs from population average
  - Relative positions, change of groups, may be misleading.

- Terminology
  - Heterogeneity
  - Unobserved heterogeneity
  - Selection
  - Selective survival
  - Other terms?
  
<!-- \begin{frame}{1. Heterogeneity at work? Black-White Crossover} -->
<!-- \end{frame} -->

<!-- \begin{frame}{1. Heterogeneity at work? Mortality Plateus} -->
<!-- \end{frame} -->


- Big Caveat: Fundamental Unidentifiability

  Same data of $N$ observations
  - $N$ draws from 1 distribution
  - $1$ draw from $N$ distributions
  - Something in-between
  

  Abel (66) and Beth (76) example. 

### A 2nd example: Exponential growth, two countries

  Two countries start equal size, but grow at different rates. What
  happens to aggregate growth rate?

```{r 2-countries, fig.cap='Aggregate growth rate of sub-populations A + B', out.width='80%', fig.align='center'}
  rA = .03 ## growth rate of A
  rB = .01 ## growth rate of A
  KA = 100 ## starting pop size
  KB = 100
  t = 0:200
  KA.t = KA*exp(rA*t) ## exp growth of A
  KB.t = KB*exp(rB*t) ## exp growth of B
  K <- KA.t + KB.t ## combined pop
  r.bar = diff(log(K)) ## growth rate
  plot(t[-1], r.bar, type = "l", ylim = c(0, 0.04),
       ylab = "r.bar", xlab = "time")
  abline(h = c(rA, rB), lty = 2)
```
<!-- title("Aggregate growth rate of sub-populations A + B") -->
- Questions

  - What determines growth rate?
  - How does it change over time?
  - Does the process converge?


### More examples to work
  1. Differential, constant mortality ($\mu_A = .03$; $\mu_B =.01$)
  2. Differential, _time-varying_ mortality or growth.
  3. ``Movers and Stayers'' (Migration)
  4. ``Movers and Stayers'' (Marriage)
  5. Fecundity: aging or heterogeneity?
  6. Divorce: duration or heterogeneity?
  7. Duration of unemployment: duration or heterogeneity?
  8. Recidivism by time out of prison

### Application

```{r echo=FALSE}
knitr::include_url('https://shiny.demog.berkeley.edu/josh/het_ruse/', height = '600px')
```
- Questions
  
  - Can you create a plateau?
  - Can you create a crossover?
  - Can you get aggregate rate to decline?
  - Anything else?
  
## Part II. Formal Analysis

### Outline
  1. Keyfitz result
  2. Keyfitz USA-Mexico example
  3. Ken's Poisson-Exponential Model
  
  
### Keyfitz result

  $${d \over dt}\bar{r}(t) = \sigma^2_r(t)$$
  
>When group-specific growth rates are constant the rate of change of the aggregate growth rate equals the variance of the growth rates.


- Derivation

By definition, 
$$ \bar{K}(t) = \sum_i K_i(t) = \sum_i K_i e^{r_i t} $$
and

$$\bar{r}(t) = {{d \over dt} \bar{K}(t) \over \bar{K}(t)}$$

Let's take derivatives and simplify, recalling definition of variance.

SOLVE?

### US-Mexico Example

```{r us-mx, fig.cap='Keyfitz result for US-Mexico', out.width='80%', fig.asp=.75, fig.align='center', fig.height=5}
rm = 3.5/100
ru = .75/100
Km = 50 
Ku = 100
t <- -50:150 ## go back in time to see rise and fall of variance
Kt = Km * exp(t*rm) + Ku * exp(t*ru)
bar.rt <- diff(log(Kt))
par(mfrow = c(2,2))
plot(t, Kt, lwd = 2, type = 'l')
title('Total pop size (solid)\n Group m (dashed)')
lines(t, Km * exp(t*rm), lty = 2, col = "red")
lines(t, Ku * exp(t*ru), lty = 2, col = "blue")
my.v = 26
abline(v = my.v)
plot(t, Kt, lwd = 2, type = 'l', log = 'y', ylim = c(.5, max(Kt)))
lines(t, Km * exp(t*rm), lty = 2, col = "red")
lines(t, Ku * exp(t*ru), lty = 2, col = "blue")
abline(v = my.v)
title("Total pop size (solid)\n Group 'm' (dashed): Log-scale")
plot(t[-1], bar.rt, type = 'l', main = 'Aggregate growth rate')
plot(t[-(1:2)], diff(bar.rt), type = 'l',
main = 'Change in aggregate growth rate')
```

#### Commentary on Keyfitz result

  - Growth rates in heterogeneous populations start at pop average
    and then increase.
  - Heterogeneity pop growth
  - We will extend to cover non-constant growth
  
  But
  
  - Doesn't tell us how much bigger $\bar{K}(t)$ is projection
    using constant aggregate rate $\bar{r}(0)$.
  - Doesn't give us a formula for time path of aggregate $\bar{K}(t)$
    or $\bar{r}(t)$
  
  Note: our homework will try to address some of this using Taylor
  approximation. 
  
### The Origin of Professor Wachter's Poisson-Exponential Model
  
Given a world with many sub-populations, each growing expontentially
at their own rate, what can we say about the time-path of world
population growth?

From an email: 

> Josh asks: Suppose we have a discrete mix of subpopulations
growing at different intrinsic rates r whose maximum is r0. Is
there a handy approximation for the growth path of the aggregate
populations?

>The assumption of a discrete mix is essential here. Otherwise
Tauberian theorems apply and, with a vanishingly small portion of
the population close to the maximum growth rate, we do not obtain
long-run exponential growth.

>I recommend modeling the discrete distribution of growth rates as
a mixture of Poisson distributions.


We are considering

\begin{equation}
  \label{mixture}
\bar{K}(t) = \sum_i e^{r_i t} K_i(0).
\end{equation}

Ken suggests 
\begin{equation}
  \label{ri_def}
r_i = r_0 - s(\lambda) \cdot a,
\end{equation}

* $r_0$ growth rate of the fastest growing sub-population
* $s$ a non-negative Poisson distributed integer
* $\lambda$ the parameter of the Poisson distribution (also
  it's mean and variance)
* $a$ gap between adjacent growth rates.


Example:  sub-populations have growth rates 3, 2, 1, 0, -1, $\ldots$ percent,
then $r_0 = 0.03$ and $a = 0.01$. Sizes of sub-pops determined by
Poisson dis'n


\begin{frame}
\begin{figure}[h]
  \caption{A simulation of heterogeneous growth}
  \centering
  \includegraphics[width = 0.8\textwidth]{../ken_model_figure.pdf}
\end{figure}
\end{frame}


#### Closed-form result

$$
K(t) = K(0) e^{r_0 t} e^{-\lambda (1 - e^{-at})}.
$$

To derive:

- Write out mixture to get 
  $$  K(t) = K(0) e^{r_0 t} \sum_i e^{-sat} f(s)
  $$ 
- Substitute for $f(s)$: $Pois ~ {\lambda^s e^{-\lambda} \over s!}$ 
- Recognize that our mixture contains the series representation of
  $e^{-at}$

Interpretation 
$$
K(t) = K(0) e^{r_0 t} e^{-\lambda (1 - e^{-at})}.
$$

- Dominant term contains the maximum population growth rate $r_0$,
- Second term gives the diminishing effect of the
  sub-populations with smaller population growth rates over time.


 Some further analysis, What is the closed-form expression for $\bar{r}(t)$?


### Some commentary


- Poisson and Exponential ``fit''
- We'll this complementarity again (e.g., with Gamma)
- Tractable  models are super powerful for enhancing our understanding.
- But be careful. Avoid extremes: the model is right/wrong.
  

A BIG caveat, are disaggregated models necessarily better?

- Some potential problems:
- Aggregate constraints?
- Interacting sub-populations?
- Illusion of precision?
  
## Conclusions
  
  - Heterogeneity as variation in risk (not just outcome)
    
  - Constantly growing parts $\neq$ constantly growth whole
    
  - Keyfitz result: Change in growth rate $=$ variance of growth
    rates
    
  - Poisson growth gives us a closed-form solution.
