# Multiplicative Fixed Frailty {#frailty}

## Outline
1. Review of Mortality Mathematics
2. Multiplicative-fixed-frailty and alternatives to it.
3. Population Survival and Hazards under fixed frailty
4. Gamma frailty

Additional resources 

  - @vaupel2014unobserved:
  - @rodriguez2001: detailed hand-out for Princeton class. Many of the same results as @vaupel2014unobserved, but with alternative derivations and notation. Also includes inversion formula and extensions beyond Gamma frailty such as “Inverse Gaussian Frailty”.

### Review of mortality mathematics

  - $\ell(x)$ or $S(x)$ probability of survival to age $x$
  - $\mu(x)$ or $h(x)$ hazard rate at age $x$ ("minus the exponential rate of change in
    survival")

  Let's treat $\mu$ as a definition.
  $$
  \mu(x) \equiv -{d \over dx} \log \ell(x)
  $$

  Can anti-differentiate (integrate) to solve for survival:
  $$
  \ell(x) = s(x) = e^{-\int_0^x \mu(a)\, da}
  $$

### Application: what is $\ell'(x)$?

  -  in words?
  -  taking derivative of $\ell(x)$
  -  interpretation

#### Two special cases
  
  -  Constant hazards $\mu(x) = \mu$. What's $\ell(x)$?
  -  Gompertz hazards $\mu(x) = a e^{b x}$. What's $\ell(x)$?
  

### Extending Keyfitz to mortality

  $$
  {d \over dx} \bar{\mu}(x) = \mbox{average rate of change} - \sigma_\mu^2
  $$

  What is $\bar{\mu}$? It's a weighted average:
  $$
  \bar{\mu}(x) = {\int \mu(x | z) \ell(x | z) p(z) \, dz
    \over
    \int \ell(x | z) p(z) \, dz}
  $$

  To derive Keyfitz extension, differentiate with respect to age
  $x$. (See \vm eq (13)). A good exercise.



Multiplicative Fixed frailty
  For individual $i$,
  $$
  \mu_i(x) = z_i \mu_0(x).
  $$
  
  - $z_i$ "frailty" of the $i$th individual. (Usually thought
    of as a random variable with mean $1$.)
  - $\mu_0(x)$ "Baseline hazard" schedule.  (Also, the
    schedule of a person with $z = 1$).
  
\begin{frame}{What are some alternatives?}

  Let's think of at least three.
  
  ($\beta$, $i$, $\Delta$)
\end{frame}

### fragile
  Which look like multiplicative fixed frailty?
```{r, fig.height = 3, echo = F, message = F, warning = F, hide = T, results = F}
  library(data.table)
  dt <- fread("/hdir/0/fmenares/Book/bookdown-master/data/SWE.Mx_5x5.txt", na.string = ".")
  dt[, x := as.numeric(gsub("-[0-9]*$", "", Age))]
  dt[Age == "110+", x := 110]
  par(mfrow = c(1,3))
  ## proportional
  dt[Year == "1990-1994", plot(x, Male, type = "l", col = "blue", log = "y", ylim = c(10^-4, 1), lty = 1)]
  dt[Year == "1990-1994", lines(x, 2 * Male, type = "l", col = "blue", lty = 3)]
  dt[Year == "1990-1994", lines(x, 0.5 * Male, type = "l", col = "blue", lty = 2)]
  title("Male (5Mx) 1990, x 0.5 and x 1.5")

  ## men and women
  dt[Year == "1990-1994", plot(x, Female, type = "l", col = "pink", log = "y",
                               ylim = c(10^-4, 1))]
  dt[Year == "1990-1994", lines(x, Male, type = "l", col = "blue")]
  title("By Sex (1990-1994)")

  ## time
  dt[Year == "1790-1794", plot(x, Male, type = "l", col = "blue", log = "y", ylim = c(10^-4, 1), lty = 3)]
  dt[Year == "1890-1894", lines(x, Male, type = "l", col = "blue", lty = 2)]
  dt[Year == "1990-1994", lines(x, Male, type = "l", col = "blue")]
  title("Male 1790, 1890, 1990")
```

## Part I. Results from Fixed Frailty


### A simulation

- Our questions
  -  How do we do a micro-simulation, with individuals?
  -  How does fixed frailty fit in?
  -  How do we compute pop survival, hazards, etc.
  -  How does life table of heterogeneous pop differ from baseline?

### Let's derive pop survival (Note: $\bar{s} = \bar{\ell}$) 

Pop survival will be a weighted average of group survival curves

  $$
  \bar{s}(x) = {p(z_1) s_1(x) + p(z_2) s_2(x) + \ldots \over
    p(z_1) + p(z_2)  + \ldots}
  $$
  With continuous $z$ (what are limits of integration?)
  
  $$
  \bar{s}(x) = \int s(x|z) p(z) \, dz
  $$

  Under Multiplicative Fixed Frailty use
  $$ \mu(x|z)  = z \mu_0(x) $$
  
  to derive
  
  $$\bar{s}(x) = \int s_0(x)^z p(z) \,dz.$$


### Now population hazards (stepping stones)

  Definition of hazards:
  
  $$
  \bar{\mu}(x) = - {d \over dx} \log \bar{s}(x)
  $$

  $$
  \bar{\mu}(x) = \mu_0(x) {\int z s_0(x)^z p(z) \, dz \over
    \int  s_0(x)^z p(z) \, dz}
  $$

  $$
  \bar{\mu}(x) = \mu_0(x) \bar{z}(x)
  $$

  Let's fill in steps.


### Rodriguez question
  
Why isn't population hazard a (simple) average of individual hazards?


> Answer: selected survival means that the distribution of frailty at 
age $x$ differs from the starting frailty distribution at age $0$.


The rate of increase in hazards

  (AKA "LAR: Lifetable Aging Rate")
  $$
  \beta(x) = {d \over dx} \log \mu(x)
  $$

  Example:

  What is $\beta(x)$ for Gompertz: $\mu(x) = a e^{bx}$?
  

Vaupel's result

  $$
  \bar{\beta}(x) = \beta_0(x) - \bar{\mu}(x) CV_z^2(x)
  $$

-  Hazards rise less slowly in pop than in baseline
-  If pop hazards plateau, then $\bar{\beta}(x) = 0$
-  Two special cases
  -  Homogeneous pop and plateau in baseline
  -  Gompertz and constant $CV_z$ (e.g., from Gamma)

## Part II. Introduction to Gamma Frailty

### The Gamma distribution
  
-  What do we want in a frailty distribution?
  -  What's the Gamma?
  -  Last math: closed form pop survival
  -  positive?
  -  a single dimension summarizing multiple factors? (Normal?)
  -  flexible?
  -  tractable?


$$
p(z | k, \lambda) = {\lambda^k  \over \Gamma(k)} z^{k-1} e^{-\lambda z} 
$$.

- $z$ the random variable representing frailty
- $k, \lambda$ parameters
- $\Gamma(k)$ A normalizing constant.

### Gamma in R

  - Mean: $k / \lambda$ 
  - Variance: $ k / \lambda^2$
  

```{r}
  ## with k and lambda
  k = 3; lambda = 6
  x <- rgamma(10000, shape = k, rate= lambda)
  mean(x)
  sd(x)
```

Alternate parameterization
```{r}
  ## with mean 1, sigma.sq
  sigma.sq <- .25
  z <- rgamma(10000, shape = 1/sigma.sq, rate = 1/sigma.sq)
  mean(z)
  var(z)
```

### Population Survival of Gamma Frailty

  Big picture
  $$
  \bar{s}(x) = \int s_0(x)^z p(z) \, dz
  $$

  Or, using our definition of survival,
  $$
  \bar{s}(x) = \int e^{-z H_0(x)} p(z) \, dz
  $$


Completing the gamma
  $$
  \bar{s}(x) = \int e^{-z H_0(x)} {\lambda^k  \over \Gamma(k)} z^{k-1} e^{-\lambda z}  \, dz
  $$

 Rearranging,
 
  $$
  \bar{s}(x) = \lambda^k \int  { 1 \over \Gamma(k)} z^{k-1}
  e^{-z (H_0(x)+\lambda)}  \, dz
  $$ 

Integral is like a $Gamma(z | k, H_0(x) + \lambda)$, but missing
something. What?

Our Result

  $$
  \bar{S}(x) = {\lambda^k \over \left[H_0(x) + \lambda\right]^k}
  $$

  If mean = 1.0, then we can let $\lambda = k = 1/\sigma^2$,
  $$
  \bar S(x) = {1/\sigma^2 \over (H_0(x) + 1/\sigma^2)^{1/ \sigma^2}} =
  {1 \over  \left(1 + \sigma^2 H_0(x)\right)^{1/ \sigma^2}}
  $$
  


### Interpreting Gamma-frailty survival
  $$
  \bar S(x) = {1 \over  \left(1 + \sigma^2 H_0(x)\right)^{1/ \sigma^2}}
  $$

  
  -  Older ages, smaller survival.
  -  Variance not so clear, need a picture. (What if $\sigma^2 = 0$?)
  
```{r, fig.height = 3}
x <- 0:100
a = 10^-4
b = 1/10
mx.0 <- a * exp(b*x)
Hx.0 <- cumsum(mx.0)
Sx.0 <- exp(-Hx.0)
## small sigma
sigma.sq = .5^2
bar.S.small.sigma <- 1 / (1 + sigma.sq *Hx.0)^(1/sigma.sq)
## big sigma
sigma.sq = 1^2
bar.S.big.sigma <- 1 / (1 + sigma.sq *Hx.0)^(1/sigma.sq)
plot(x, Sx.0, lty = 2, type = "l", ylim = c(0,1),
       ylab = "Survival")
  lines(x, bar.S.small.sigma, col = "blue")
  lines(x, bar.S.big.sigma, col = "red")
  legend("bottomleft", c("Pop big.sigma",
                       "Pop small.sigma",
                       "Baseline"),
         lty = c(1, 1, 2),
         bty = "n",
         col = c("red", "blue", "black"))
  title("Gamma-frailty population survival")
```

## Conclusions
  - Multiplicative Fixed Frailty is one option for modeling
  - Gave us analytical expressions for population survival and
    hazards including $\bar{\mu}(x) = \mu_0(x) \bar{z}(x)$
  - Extended Keyfitz result to age-changing hazards
  - Survival curve for Gamma
  
