# Convergence and cross-overs

## Outline
- Concepts
- Student Presentation

Additional resources:   

  - @coale1986mortality: Non-technical overview of data issues that could create the appearance of mor-tality cross-overs.  
  - @manton1981methods: interesting discussion on parameter choice of the Gamma and the empirical results.  


## What happens to mortality disparities at older ages?
  
  - Cumulative disadvantage
  - Age as a leveler
    - Individual adaptation/plasticity, gov support, separation from
    unequal structures like labor market
  - Bad data / measurement
    - Unreliable ages, institutionalization changes sample, etc.

  - Nothing
    - It's all selection ("frailty"), pop hazards but
    individual hazards would have remained "parallel".

Our goal is to examine this last "null hypothesis". What can  
  frailty explain, and what can't it?
  
### A possible null-model

- 2 groups, each with internal gamma-frailty
- proportional baseline hazards

  $$
  \mu_2(x) =  R \mu_1(x)
  $$
see V&M (38)

  $$
  \mu_1(x |z_1) = \mu_1 z_1 \\ \mu_2(x |z_2) = \mu_2 z_2  
  $$

  And, frailty terms are each gamma, with mean 1 and own variances.

### A result: \vr (5E)

  $$
  \bar{R}(x) \equiv {\bar\mu_2(x) \over \bar\mu_1(x)} =
  {R + R\sigma_1^2 H_1(x) \over 1 + R \sigma_2^2 H_1(x)}
  $$
  
  Questions:
  
  \item If variances are equal. What happens at age 0? What happens
    at very old ages.
  \item If the higher mortality group has bigger frailty variance, what
    happens at older ages?
  \item Same if  higher mortality group has smaller frailty
    variance?
  

> Homework: prove this, simulate this. see if cross of is when
  cumulative hazards satisfy the condition at the end of 5E). (*
  problem. can you solve for x0 in temrs of variances 1 and 2 and R
with gamma gompertz?

### Inversion

Our challenge is to invert a not easy pop hazards formula
$$
\bar{\mu}(x) = {\mu_0(x) \over 1 + \sigma^2 H_0(x)}
$$
because we have both hazards and cumulative hazards on right.

Hazards are slope of log survival

Recall for Gamma,
$$
\bar{S}(x) = { 1 \over (1 + \sigma^2 H_0(x))^{1/\sigma^2}}
$$

We write down the hazard as the derivative of log survival
$$
\bar{\mu}(x)  = {1 \over \sigma^2} {d \over dx} \log(1 + \sigma^2 H_0(x)).
$$

The anti-derivative of both sides, gives
$$
\bar{H}(x) = {1 \over \sigma^2} \log(1 + \sigma^2 H_0(x)).
$$

And now we have only 1 expression involving the baseline hazards on
the right. 


Solving 
$$
\bar{H}(x) = {1 \over \sigma^2} \log(1 + \sigma^2 H_0(x)).
$$
gives us the cumulative hazard
$$
H_0(x) = {1 \over \sigma^2} \left(e^{\sigma^2 \bar{H}(x)} - 1 \right).
$$

And differencing, gives us a remarkably simple expression for the
baseline hazard in terms of the observed popualtion hazard
$$
\mu_0(x) = \bar\mu(x) e^{\sigma_2 \bar{H}(x)}
$$

- We don't observe underlying baseline hazard $\mu_0$ on left
- What is observed (and unobserved) on right?


### An example

```{r} 
library(data.table)
dt <- fread("/hdir/0/fmenares/Book/bookdown-master/data/ITA.cMx_1x1.txt",
            na.string = ".")
my.dt <- dt[Year == 1915]
my.dt[, H.f := cumsum(Female)]
my.dt[, H.m := cumsum(Male)]
my.dt[, h.f := Female]
my.dt[, h.m := Male]
sigma.sq <- .5^2
my.dt[, h0.f.5 := h.f * exp(sigma.sq *H.f)]
my.dt[, h0.m.5 := h.m * exp(sigma.sq *H.m)]
sigma.sq <- .2^2
my.dt[, h0.f.2 := h.f * exp(sigma.sq *H.f)]
my.dt[, h0.m.2 := h.m * exp(sigma.sq *H.m)]
sigma.sq <- 1^2
my.dt[, h0.f1 := h.f * exp(sigma.sq *H.f)]
my.dt[, h0.m1 := h.m * exp(sigma.sq *H.m)]
```


Italian Females, born 1915 ($\sigma^2 = .2^2, .5^2, 1^2$)

```{r fig.height=6, warning=FALSE}
par(mfrow = c(1,2))
foo <- my.dt[, plot(Age, H.f, col = "red")]
title("Cumulative Hazards\n Italian Females born 1915")
foo <- my.dt[, plot(Age, log(h.f), type = "l", ylim = c(-7, 2), col = "red",
                    main = "Observed vs. implied baseline")]
foo <- my.dt[, lines(Age, log(h0.f.2), lty = 2, col = "red")]
foo <- my.dt[, lines(Age, log(h0.f.5), lty = 3, col = "red")]
foo <- my.dt[, lines(Age, log(h0.f1), lty = 4, col = "red")]
legend("topleft", legend = c("baseline", "obs if s2 = .2^2", "obs if s2 = .5^2", "obs if s2 = 1^2"),
       col = "red", lty = 1:4)
```



Italian Females vs Males born 1915 ($\sigma^2 = .5^2$)
```{r fig.height=6, warning=FALSE}
par(mfrow = c(1,2))
## title("Cumulative Hazards\n Italian Females born 1915")
foo <- my.dt[, plot(Age, log(h.f), type = "l", ylim = c(-7, 2), col = "red",
                    main = "Observed vs. implied baseline")]
foo <- my.dt[, lines(Age, log(h0.f.5), lty = 2, col = "red")]
ugh <- my.dt[, lines(Age, log(h.m), type = "l", col = "blue")]
ugh <- my.dt[, lines(Age, log(h0.m.5), lty = 2, col = "blue")]
legend("topleft", legend = c("female observed",
                             "female baseline",
                             "male observed",
                             "male baseline"),
       col = c("red","red", "blue", "blue"), lty = c(1,2, 1,2))
## now do difference
foo <- my.dt[, plot(Age, h.m/h.f, type = "l", col = "black",
                    ylab = c("h.m/h.f"),
                    main = "Male-female hazard ratio")]
foo <- my.dt[, lines(Age, h0.m.5/h0.f.5, lty = 2)]
legend("topright", legend = c("observed",
                              "implied baseline"),
##       col = c("red","blue"),
       lty = 1:2)
```

Much bigger convergence in ``observed'' than in baseline

### Application

```{r echo=FALSE}
knitr::include_url('https://grodri.shinyapps.io/heterogeneity/', height = '600px')
```

## Student Presentation
  