# Tempo

## Outline

- A tempo simulation
- Bongaarts and Feeney's formula
- An application to the United States
- Two Americas?
- EM algorithm for unmixing mixtures
- An application to two Americas.

## Summary 

### What we see is superficial. Heterogeneous models reveal what's ``really'' going on. (Or do they?)
  
- Until today, population hazards mislead
- Today, homogeneous fertility misleads

### Wwill reverse perspectives

  - We see differences we see in genotypes, in lineages, in
    names.
  - These could be due to ``real'' differences
    (heterogeneity).
  - But they could also be due to luck. Everyone is the same
    but stochastic outcomes differ.
  - Our models of individual-level randomness will have
    predicted dynamics, which are themselves interesting but can
    also be used as a ``null'' to compare to observations.


## Fertility postponement, a very simple example

  Baseline
  
  - A population has a history of 1 birth per day
  - When women turn age 25, they have a daughter.
  - This gives us a constant stream of births, 365 per year.
  

  Postponement
  
  - Starting on Jan. 1, 020, everyone postponements childbearing
    an additional month, until they are aged 25 1/12.

  -  How many births will there be in 2020?

  -  How many births in 2021?
  
### Continuous postponement, a shiny simulation
  
```{r echo=FALSE}
knitr::include_url('https://shiny.demog.berkeley.edu/josh/tempo/', height = '600px')
```

  - [$R(t)$] Cumulative postponment
  -[$r(t)$] Incremental postponement $r(t) = R'(t)$


  What is a formula for recovering original birth stream?
  $$
  \hat{B}_{orig} = B_{obs} \times (1 + R'(t))
  $$
  or
  $$
  \hat{B}_{orig} = B_{obs} \times 1/  \left[1 - R'(t)\right]?
  $$

  Note: this idea of ``recovering original'' is one way to think
  about tempo adjustment.
  
A bigger microsimulation
  
  - Each period will have births across a range of ages
  - We'll randomly generate the original planned birthdays
  - Then we'll shift by a continuous function $R(t)$.
  

## Period Shifts: Bongaarts and Feeney's model

  $$
  f(a,t) = f_0(a - R(t)) (1- R'(t)) q(t)
  $$
  
  
  -[$f(a,t)$] birth rate of women aged $a$ in period $t$
  -[$f_0$] A constant baseline schedule (can be norm'd to sum to 1).
  -[$q(t)$] A period intensity parameter: ``quantum''
  -[$R(t)$] Cumulative shift.
  

### An example

  $$
  f(a,t) = f_0(a - R(t)) (1- R'(t)) q(t)
  $$
  
  
  - $R_{2019} = 3$
  - $R'_{2019} = .1$
  - $q(2019) = 1$
  

  Give an expression for $f(28,2019)$.


## A derivation: due to Rodriguez

  Assume no quantum effects.

  Take a cohort with cumulative fertility
  $$
  F_0(a) = \int_0^a f(x) \,dx
  $$

  Now put in shifts so that observed fertility is from an
  age $R(t)$ years \alert{earlier}. (``28'' is the new ``25''!)
  $$
  F(a,t) = F_0(a - R(t)) \pause = F_0(a - R(c + a))
  $$
  
  Differentiate with respect to age (which for a cohort is also
  time $t$), using chain rule
  
  $$
  f(a,t) = f_0(a - R(t)) \left[1 - R'(t)\right]
  $$
  Bingo!


### Quantum comes at the end
  Let's re-notate our constant quantum result
  $$
  f_0(a,t | R(t) ) = f_0(a - R(t)) \left[1 - R'(t)\right]
  $$
  Then we can incorporate period quantum on the shifted surface:
  $$
  f(a,t) = f_0(a,t | R(t) ) q(t) = f_0(a - R(t)) \left[1 - R'(t)\right]q(t)
  $$

  Note: If we vary quantum \bold{before} shifts, then $q(t)$ will
  bleed into neighboring years. (a small effect, but makes model
  messier).


### Tempo-adjusted TFR: counter-factual, TFR in absence of timing changes

  $$
  TFR(t) = \int_0^\infty f(a,t) \, da
  $$

  Substituting our shifted birth rates with quantum
  $$
  TFR(t) = \int_0^\infty f_0(a - R(t)) \left[1 - R'(t)\right]q(t) 
  $$
  gives?
  \pause

  
  $$
  TFR(t) = TFR_0 \left[1 - R'(t)\right] q(t)
  $$
  \pause
  WLG, define $TFR_0 = 1$, then
  $$
  q(t) =   \alert{ {TFR(t) \over 1 - R'(t)} \equiv TFR^*(t) }
  $$
  Voila, the BF formula


### How do period schedules change?
  For homework
  $$
  f(a,t) = f_0(a - R(t)) \left[1 - R'(t)\right]
  $$


  What does
  $$
  {\partial \over \partial t} \log f(a,t) = ?
  $$
  Let's sketch

## ``Uniform'' shifts

  - BF model assumes all ages shift by $R(t)$.
  - BF model assumes all ages rise or fall by same quantum $q(t)$
  - Violating these assumptions means change in mean age will not
    just reflect ``tempo''.
  - Example: What happens if people have fewer higher order births?
  
## BF recommendation for achieving uniformity

  Separate estimates for each birth order, and then combine:
  $$
  TFR^*(t) = \sum_i TFR_i^*(t) = \sum_i {TFR_i(t) \over 1 - r_i(t)}
  $$

  This will protect against order-specific quantum effects.
  

## An Application to the United States

  - We'll use HFD data
  - We'll do tempo adjustment for all births
  - We'll redo by birth order

```{r}
## Tempo adjustment of US fertility using HMD data and
## Bongaarts-Feeney formula

library(data.table)
source("/hdir/0/fmenares/Book/bookdown-master/codes/tempo_functions.R")
source("/hdir/0/fmenares/Book/bookdown-master/codes/utility_functions.R")


## (1) read in data and format into an array
## (2) fit bongaarts feeney without birth order
## (3) fit bongaarts feeney with birth order

## (1) read in data and format into an array

## age specific fertility rates by birth order for all countries and times
## RR means "rectangles" on Lexis surface
dt <- fread("/hdir/0/fmenares/Book/bookdown-master/data/zip_w/asfrRRbo.txt")
dt <- dt[Code == "USA"] ## keep only US
## keep only ages 15 to 49
dt <- dt[Age %in% 15:49]
print(dt)

## put all order fertiility into a matrix

fat <- dt[, xtabs(ASFR ~ Age + Year)]
fat <- as.matrix(unclass(fat)) ## may cause problems, we'll see

fat1 <- dt[, xtabs(ASFR1 ~ Age + Year)]
fat2 <- dt[, xtabs(ASFR2 ~ Age + Year)]
fat3 <- dt[, xtabs(ASFR3 ~ Age + Year)]
fat4 <- dt[, xtabs(ASFR4 ~ Age + Year)]
fat5p <- dt[, xtabs(ASFR5p ~ Age + Year)]

year.vec <- colnames(fat)
age.vec <- rownames(fat)
parity.vec <- c("all", 1:5)
fat.array <- array(NA, dim = c(nrow(fat), ncol(fat), length(parity.vec)))
dimnames(fat.array) <- list(age.vec, year.vec, parity.vec)
fat.array[,,"all"] <- fat
fat.array[,,"1"] <- fat1
fat.array[,,"2"] <- fat2
fat.array[,,"3"] <- fat3
fat.array[,,"4"] <- fat4
fat.array[,,"5"] <- fat5p

## (2) fit bongaarts feeney

tfr.vec <- colSums(fat)

## (2a) by hand
mu.vec <- apply(fat, 2, get.mean)
rt.vec <- center.diff(mu.vec)
adj.tfr.vec <- tfr.vec / (1 - rt.vec)

par(mfrow = c(3,1))
plot(names(mu.vec), mu.vec)
plot(names(mu.vec), rt.vec)
abline(h =0)
plot(year.vec, tfr.vec, type = "l")
lines(year.vec, adj.tfr.vec, lty = 2)
abline(v = c(1945, 2008))
## we see fertility since 1980 has been depressed by postponment
## we see weird dynamics around end of WW2 and great recession.
## what's going on?

par(mfrow = c(1,1))
plot(year.vec, tfr.vec, type = "l")
lines(year.vec, adj.tfr.vec, lty = 2)
abline(v = c(1945, 2008))
abline(h = seq(1.5, 2.5, .1), col = "grey", lty = 3)

## can also use function to fit

adj.tfr.vec.from.fun <- bf.fit.simple(fat)$tfr.star
lines(year.vec, adj.tfr.vec.from.fun, col = "red")

## Now let's look at turbulence around WW2
par(mfrow = c(2,2))
plot(age.vec, fat[,"1944"], type = "l", ylim = c(0, .23),
     ylab = "f(a)",
     xlab = "age a"
     )
lines(age.vec, fat[,"1945"], type = "l", col = "red")
lines(age.vec, fat[,"1946"], type = "l", col = 'orange')
lines(age.vec, fat[,"1947"], type = "l", col = "blue")
legend("topright",
       legend = 1944:1947,
       col = c("black", "red", "orange", "blue"),
       lty = 1)
title("Age specific fertility")
##
plot(1943:1947, mu.vec[paste(1943:1947)],
     ylab = "mu(t)",
     xlab = "t",
     col = c("black", "black", "red", "orange", "blue"),
     pch = 19)
title("Mean ages")
##
plot(1943:1947, rt.vec[paste(1943:1947)],
     ylab = "r(t)",
     xlab = "t",
     col = c("black", "black", "red", "orange", "blue"),
     pch = 19)
title("Changes in mean, centered")
##
plot(1943:1947, tfr.vec[paste(1943:1947)],
     ylab = "tfr",
     xlab = "t",
     ylim = c(1, 4),
     type = "l")
lines(1943:1947, adj.tfr.vec[paste(1943:1947)],
      lty = 2)
title("TFR and adjTFR")

## From 1945 to 1946, fertility goes up a lot, but more at younger
## ages. So mean goes down. BF adjustment over-compensates, and has
## quantum declining.

## What's happening from 1944-45?


## (3) fit bongaarts feeney with birth order

out <- bf.fit(fat.array)

adj.tfr.bo.vec <-  out$tfr.star.out[, "bf.tfr.star"]

## pdf("usa_tempo_fig.pdf")
par(mfrow = c(1,1))
plot(year.vec, tfr.vec, type = "l", lwd = 2)
lines(year.vec, adj.tfr.vec, lty = 2)
lines(year.vec, adj.tfr.bo.vec, lty = 1, lwd = 2, col = "red")
## let's check against hfd
hfd.adj.dt <- fread("/hdir/0/fmenares/Book/bookdown-master/data/zip_w/adjtfrRR.txt",
                skip = 2)
hfd.adj.dt <- hfd.adj.dt[Code == "USA"]
hfd.adj.dt[, points(Year, adjTFR, col = "red", cex = .8)]
legend("topright",
       c("tfr", "tfr* (all parities)", "tfr* (by parity)"),
       col = c("black", "black", "red"),
       lty = c(1, 2, 1),
       lwd = c(2,1,2))
hfd.adj.dt[, lines(Year, filter(adjTFR, rep(1/7, 7)), col = "red", lwd = 4)]

## dev.off()

## Q. Does taking birth order into account smooth WW2 turbulence?
## Q. Does taking birth order into account smooth Recession turbulence?
## Q. Does taking birth order into account, retell the baby boom?
## Q. Does taking birth order into account, retell the baby bust?


```




  - Baby boom smaller if we account for ``pre-ponement''.
  - Fertility lull in 1970s and 80s disappears if we account for
    ``postponement''
  - Birth order disaggregation improves estimates of shifts from
    changes in mean age
  - What happened with the recession?

## Two Americas

  Let's look at births (all orders).


- fat_movie.pdf

  Let's look at 1st births, again as if their are two latent groups:
  $A$ and $B$.  (These could be ``early moms'' / ``late moms'',
  non-college / college,
  pre-marital / marital, lower-class / upper class, $\ldots$)


- fat_mix_movie.pdf

```{r}
  ## simulate 2 normals 
  N <- 1000
  x1 <- rnorm(N, mean = 22, sd = 3) ##
  x2 <- rnorm(2*N, mean = 30, sd = 4)
  ## combine them
  x <- c(x1,x2)
  ## use EM to infer mixture
  out <- normalmixEM(x,
                     lambda = c(.5, .5),
                     mu = c(15, 35),
                     sigma = c(5,5))

  print(out$mu)
  print(out$sigma)
  print(out$lambda)
```
  Seems to work great.

```{r, fig.height = 4}
  ages <- 10:49
  dens1 <- dnorm(x = ages, mean = out$mu[1], sd = out$sigma[1]) * out$lambda[1]
  dens2 <- dnorm(x = ages, mean = out$mu[2], sd = out$sigma[2]) * out$lambda[2]
  par(mfrow = c(1,1))
  hist(x, probability = T, col = "grey")
  lines(ages, dens1, col = "red", lwd = 2)
  lines(ages, dens2, col = "blue", lwd = 2)
  lines(ages, dens1 + dens2, col = "black", lwd = 2)
``` 

### An algorithm for tempo adjustment of mixtures
  
  - Fit normal mixture to each year.
  - Refit using constant variance (average). This assures shape
    invariance of each component, fulfilling BF assumption.
  - Estimate BF separately for $A$ and $B$, and combine.



- tempo_mixed_results_figure.pdf
  

Identifiability?

% could do with 3 sub-populations?

## Conclusions 
  
  - Postponement dilutes period births, lowers TFR
  - Tempo-adjustment tries to ``put births back in''
  - Changes in mean work fine if ``shape'' doesn't change
  - Shape can change through heterogeneity
  - With strong assumptions, we can identify heterogeneity
  - Declining quantum for young and postponement for old appears
    to be the story
  
### Caveats
  
  - Who are these latent groups? Do you start out in one and end
    up in the other? Do you stay in one your whole life?
  - How do we project forward?
  - Can we use other indicators (e.g., social class, education,
    marriage) to get same results?








