# Fisher-Wright

## Outline
-Example

## Parallel


<style>
.column-left{
  float: left;
  width: 50%;
  text-align: left;
}
.column-right{
  float: right;
  width: 50%;
  text-align: left;
}
</style>


<div class="column-left">
*Fisher-Wright*

  - Children picking their parents (not ``generative'')
  - Total population size is constant
  - Qualitatively similar to BP. Extinction and fixation.
  - Flexible: mutation, selection, even changes in pop size.
  - With apologies, biologists take FW ``seriously'' even if they
    don't take it ``literally.''
    
</div>
<div class="column-right">
  *Galton-Watson-Bienaym\'e Branching Processe*
  - Branching process models independent parents randomly
    producing offspring. ``Generative''
  - Total population size can vary, and has a random component and
    deterministic one $m$
  - Qualitative result when $m = 1$ is that there is one longest
    surviving line. This is ``fixation'', when one {\em type} becomes
    universal.
</div>
### Another cell phone example
Let's simulate and draw lines
Gen 1   0    1    2    3    4    5    6    7    8    9
Gen 2
Gen 3
- What will happen?
Simulation code
```{r}
  fwm <- function(N, n_gen, mu = 0) ## mu != 4/N
  {
      ## simulate fisher-wright (with mutations)
      x <- paste(1:N) ## starting types
      A <- matrix(NA, nrow = n_gen, ncol = N)
      for (i in 1:n_gen)
      {
          A[i,] <- x
          x <- sample(x, size = N, replace = T)
          x <- mut(x, mu)
          x
      }
      return(A) ## matrix of types, each line a generation.
  }
```


```{r}
  mut <- function(x, mu)
  {
      ## m, the individuals that mutate
      m <- which(rbinom(length(x), 1, mu) == 1)
      if (length(m) == 0) ## if no-one mutates
          return(x)
      ## add a suffix to their ID, so it will be unique (infinite alleles)
      suffix <- 10000*round(runif(length(m)),4)
      x[m] <- paste0(x[m], ".", suffix)
      x
  }
```

Trying it out
```{r}
  set.seed(1)
  fwm(N = 10, n_gen = 2, mu = 0)
  fwm(N = 10, n_gen = 2, mu = 0)
```   

Trying it out some more

```{r, fig.height = 3.5}
  set.seed(1)
  A <- fwm(N = 10, n_gen = 20, mu = 0)
  tt <- table(A, row(A)) ## count types by row
  ptt <- prop.table(tt, 2) ## proportions
  matplot(t(ptt), type = 'l', lty = 1, main = "FW simu")
  text(x = 4, y = jitter(ptt[,4]), rownames(ptt), col = 1:6)
``` 

  Qs: What happens at time 15?  Why does line 5 rise and fall? What happens at time 2?
  What is $E(p_i(t) | p_i(t-1))$?


Bigger pop and more time
```{r, fig.height = 4}
  set.seed(1)
  A <- fwm(N = 100, n_gen = 200, mu = 0)
  tt <- table(A, row(A)) ## count types by row
  ptt <- prop.table(tt, 2) ## proportions
  matplot(t(ptt), type = 'l', lty = 1)
``` 
  
  - What does this remind you of? What will happen in long run?
  - What other questions could we ask?
  

## Questions we can ask
  
  - What is probability that line $i$ will ``fix''? (Hint: easy)
  - What is expected time until some line fixes? (We'll demo the result)
  - How can we describe the path to fixation? (We'll derive the result)
  
  
### Probability that a particular line will "fix"
```{r, fig.height = 5}
  set.seed(1)
  A <- fwm(N = 10, n_gen = 20, mu = 0)
  tt <- table(A, row(A)) ## count types by row
  ptt <- prop.table(tt, 2) ## proportions
  matplot(t(ptt), type = 'l', lty = 1, main = "FW simu")
  text(x = 4, y = jitter(ptt[,4]), rownames(ptt), col = 1:6)
``` 
### Expected time until fixation?
  Answer for us is
  $$
  \bar{T}_{fixed} = 2 \cdot N
  $$
  Note: Biologists say $4 N_e$ \\
  See Wikipedia "Genetic drift"
### Simulation of time to fixation
```{r}
  T.vec <- NULL
  all.the.same <- function(x){length(unique(x)) == 1}
  set.seed(10)
  for (i in 1:100)
  {
      A <- fwm(N = 100, n_gen = 1000,mu = 0)
      extinction_time = min(row(A)[apply(A, 1, all.the.same)])
      T.vec[i] <- extinction_time
  }
  mean(T.vec)
```
## Path to fixation: a measure of homogeneity/heterogeneity

  Chance that two randomly drawn individuals are of same type.
  $$
  G = \sum_i p_i^2
  $$

  If we have two types, $p_1 = \pi$ , $p_2 = 1-\pi$?
  What is G if $\pi = 0, .5, 1$?


Let's derive time path of G

  Let's assume just two types, $\pi(t)$

  Chance two indiv are of same type
  $$
  G_{t+1} = P(\mbox{same parent})\cdot 1 +
  P(\mbox{different parent})\cdot G_{t}
  $$

  Notation: I'm going use $K$ for pop size. Bio uses $2N$.
  $$
  G_{t+1} = {1 \over K} \cdot 1 + (1 - {1\over K}) \cdot G_{t}
  $$

  Easier to solve letting $H = 1 - G$. Some algebra gives
  $$
  H_{t+1} = H_{t} (1 - 1/K)
  $$

  Or,
  $$
  H_{t} = H_0 (1 - 1/K)^t % \rightarrow H_0 e^{-t/K}
  $$
  So, H goes to 0 exponentially, just as G goes to 1.


## Baby Names

  "Drift as a mechanism for cultural change: an example from baby names"
  by Matthew W. Hahn and R. Alexander Bentley
>  Proc. R. Soc. Lond. B 2003 270, S120-S123

What's the basic idea?
  
  - How is naming a baby like Fisher-Wright?
  - How is it not?
  

  - Like Fisher-Wright 
    - people choose from existing set (?)
    - names are ``neutral'' (?)
    - draw proportionally (?)
  - They test to see if they can reject FW
    - compare observed histograms to FW simulation
  - They include mutation to get a stationary disn
  Note: failing to reject FW doesn't mean it's correct
Their picture 
>  \includegraphics[width=0.4\textwidth]{./hahn_pic_1.png}
> \includegraphics[width=0.6\textwidth]{./hahn_pic_2.png}
### Drawing their picture with the data
  babynames.R
### Some potential criticism
  
  - While we can't reject that there's some parameterization of
    FW that gives us similar disn, this doesn't mean that we've
    found the right mechanism. (Just that we can't reject it).
  
  - What are some other tests of this mechanism?
  
  - Markov assumption. We could see if each frequency really
    followed random walk.
  - Perhaps we could see if variances were scaled to
    frequencies correctly.
  
## Conclusions
  
  - Fisher-Wright an alternative to branching processes
  - It reverses logic of reproduction, but gives similar
    quantitative and qualitative results
  - A neutral model for other processes?
  - Starting point for coalescent