# Coalescent 

\title{ \\ Dem260 Math Demog

## Outline   
  
  \item Big picture: What is ``coalescent theory''?
  \item Time to (T)MRCA
  \item Simulation: Inferring population size
  \item An application of Coalescent Theory
  

## Big picture: The Coalescent: Expectations of the Past
  
  \item Coalescent theory is not a theory.
  \item It's a model for the probability of different histories
  \item ``the'' coalescent is a bit confusing. We're not inferring the
    actual history of common ancestry, just the probabilities
 
An actual``picture''
> includegraphics[width=1.1\textwidth]{../coalescent/cutter_diagram.png}
     
\item Top panel is a Fisher-Wright instance, ordered so that lines
         don't cross.
\item Haplotype is a sequence (we are diploids, each 
         contributing 2 haplotypes). But let's just think of each line as
         an individual, for now.
\item We can find The Most Recent Common Ancestor (TMRCA) of 
sample (dark purple). Who and when would the MRCA of the top two individuals be?


  \item Our sample $\neq$ even all {\em extant } descendants of the MRCA. What
    does this mean?
  \item Our sample $\neq$ all of the descendants of the MRCA. What
    does this mean?
  \item If we chose two descendants at random, would we always get
    same MRCA? 
  
When we model coalescence we are thinking backwards in time.
  
## Our first question: When was MRCA?

  If we sample two individuals (today), how long ago was their MRCA?\\
  (Note: question is not ``who'')
  
  \item Our answer will in terms of the probability of MRCA being 1
    generation ago, 2 generations ago, etc.
  \item We'll assume Fisher-Wright (constant N, each gen randomly
    picks parents)
  \item The answer is surprisingly simple
  
Let's assume we have $N$ lines in Fisher-Wright (Note: I'm not using $2N$.)
  
  \item The chance that two sampled people have same parent is $1/N$, right?
  \item Thus $P(T_{MRCA} = 1) = 1/N$.
  \item What is $P(T_{MRCA} = 2) = $?
  \item What is $P(T_{MRCA} = n) = $?
  
  Let's go to continuous time (reasonable if pop is big and time scale
  is long).
  
  Hazard of coalescence = $c = 1/N$.
  Probability of coalescence at time $t$ = $\ell(t) h(t) = e^{-ct} c$

  What is expected time of coalescence? Think life expectancy.
  
  $E(T_{MRCA})$ if two samples: $1/c = 1/(1/N) = N$

  Let's simulate

  \item 1 time, without random seed, letting N = 40, ngen = 200, mu = 0
  \item Average over 100 FW simulations
  \item What is variance of outcome? Is it what we would expect from
    exponential?


## Mutation and inference of TMRCA and $N$
  
  \item Say mutations occur at a constant rate $\mu$ ($10^{-8}$?)
  \item Each year we would expect $\mu$ mutations, and over $T$ years we
    would expect $T\mu$ mutations.
  \item Say we observe that two people differ at $k$ sites of the
    genome.
    \item When was TMRCA?
    \item How big is the population?
  


  Picture ($\Lambda$)
  Tree length = $2T$
  Expected number of mutations:
  $$
  E(k) = E(2T\mu) = \bar{T} 2 \mu
  $$
  Since,
  $$
  \bar{T} = E(TMRCA) = N
  $$
  If we observe on average $\bar{k}$ mutations, then
  $$
  E(k) = N 2 \mu \rightarrow \hat{N} = {\bar{k} \over 2 \mu}
  $$ 

## Inference of population size, simulation
  
  \item We do FW with mutations
  \item Average pairwise differences
  \item Divide by $2\mu$ to get our estimate
  \item We can repeat a bunch of times and see average estimate
    converges to the truth
  

Coalescence of a sample of $n$ individuals
  
  \item This is covered on pages 42 and 43 of Gillespie
  \item We'll just do one quick example, accepting the result
  
A sample of 3: Note we're using $N$ (instead of $2N$)

\begin{verbatim}

        *            _______
       / \
      /   \          
     /     \         T(2) : E(T(2)) = N
    /       \
   /         *       _______
  /         / \
 /         /   \     T(3) : E(T(3)) = N * 2/[3*2]
*         *     *    _______

                      ...
                     _______
                     T(n) : E(T(n)) = N * 2/[n * (n-1)]
\end{verbatim}
  Question: If we sample 4, how much of time to TMRCA
  is do we have 4 branches, 3 branches, and 2 branches?

## Conclusions


  \item We defined the coalescent as the stochastic process going
    back in time to common ancestors
  \item For constant population size, we proved that time to
    coalescence for a sample pair is exponential.
  \item We showed (math and simulation) that $E(T) = N$.
  \item We showed that we could estimate $N$ from observed mutations
    if we knew the mutation rate


\end{frame}

## An application of coalescent theory
  
Application to making inferences about the real history of human populations.
    

 
Caveat: We're still using $N$ (not $2N$) as the number of haploids

### Coalesence when population is changing
  
  \item Ee said hazard of coalescence was $h = c = 1/N$.
  \item What is hazard of coalesence in one generation for two
    different populations: $N =
    1000$? $N = 2000$?
    
  \item What if within the same population $N(t) = 1000$ and $N(t+1)
    = 2000$? (Hint: we still follow FW in allowing children to
    choose their parents.)
    
  \item If the population size changes over time $N(t)$, then hazards
    of coalescence in will change too: $h(t) = 1/N(t)$.
  
```{r ,fig.height = 3}
  N_recent = 5000 ## population last T_thresh years
  T_thresh = 1000
  N_ancient = 500 ## earlier population
  n = 1000 ## sampled individuals
  set.seed(0.4886)
  T1 <- rexp(n, rate = 1/N_recent) ## give everyone a chance to coalesce 
  T1[T1 > T_thresh] <- NA ## if they don't in 1st 1000 years, resample them
  n2 <- sum(is.na(T1))
  T2 <- T_thresh + rexp(n2, rate = 1/N_ancient) ## at ancient rate
  T.vec <- c(T1, T2)
  hist(T.vec, breaks = seq(0, 5000, 250))
```
  Q: How could we estimate population sizes from this histogram?
  
```{r, fig.height = 3}
  T.vec <- sort(T.vec)
  St = (n:1)/n
  par(mfrow = c(1,2))
  plot(T.vec, St); abline(v = T_thresh); plot(T.vec, log(St)); abline(v = T_thresh)
```
  Q: How can we estimate hazards from this histogram?



Our approach
  
  \item Say we have $i$ pairs of haploids
  \item We then compute how many pairwise differences there are,
    but instead of computing $\bar{k}$, we keep the distributional
    information $k_i$.
  \item Each $k_i$ implies a $T_i$
  \item We then have a set of ``death times'' (coalescence times),
    can build a life table, estimate the hazards, and infer
    $N(t)$.

### Inference from MRCA times: $n > 2$ comparisons and relative branch lengths
    
Detour to length of branches
A sample of 3: Note we're using $N$ (not $2N$)
\begin{verbatim}

        *            _______
       / \
      /   \          
     /     \         T(2) : E(T(2)) = N
    /       \
   /         *       _______
  /         / \
 /         /   \     T(3) : E(T(3)) = N * 2/[3*2]
*         *     *    _______

                      ...
                     _______
                     T(n) : E(T(n)) = N * 2/[n * (n-1)]
\end{verbatim}

  Intuition: when we have more individuals, there's more chance that
  some pair of them will coalesce.
  
  Question: If we sample 4, how much of time to TMRCA
   do we have 4 branches, 3 branches, and 2 branches?


## Using real mtDNA to estimate population sizes. 
  Let's go to .Rmd in RStudio







