---
title: "Simulation of rates of mortality improvement by age in hetogeneous cohorts"
author: Joshua R. Goldstein
output: html_notebook
---

# Overview

We want to see what it means for frailty to "slow mortality
progress". To do this we simulate two life tables, one with 10% lower
mortality at all ages and show that at older ages the gap is much
smaller.

(A more general example could have different decreases at different
ages.)

We show that the simulation produces what should be expected from the
gamma-gompertz.

The result has a kind of "duality"

* On the one hand it is repeating what we did with cross-overs,
  showing that differences between observed hazards of groups is
  biased downward, when compared to differences in base-line hazards.
  
* On the other hand, if we can conceive of these groups as different
  cohorts on the Lexis diagram, giving us the interpretation that
  slower progress at older ages may also be driven by mortality
  seleciton.
  

## Preliminaries

```{r}
source("../gomp_funs.R")
get.Dx <- function(y)
{
    ## counts number of x in single year age groups
    ## including zeros when there's no one
    ## (note: built-in "table()" won't do this :(
    x <- 0:max(floor(y))
    y.fac <- factor(floor(y), levels = x)
    Dx <- tabulate(y.fac)
    names(Dx) <- x
    return(Dx)
}
get.lt <- function(y)
{
    Dx <- get.Dx(y)
    x <- as.numeric(names(Dx))
    ## get lx by reverse-survival
    lx <- rev(cumsum(rev(Dx)))
    ## get person-years as average of adjacent lx
    lxpn <- c(lx[-1], 0)
    Lx <- (lx + lxpn)/2
    ## get hazards
    mx <- Dx/Lx
    Sx <- lx/lx[1]
    lt <- data.frame(x=x, Sx=Sx, mx=mx)
    return(lt)
    }
```


## Simulate our frailty (z's)

We do one set of z's, which we are going to use twice. (It would also
be fine to do two sets of z's, but one set runs faster with big sample.)

```{r}
million = 10^6
N <-  4*million
my.sd <- sqrt(.2) ## .44
sigma.sq <- my.sd^2
z <- rgamma(N, shape = 1/sigma.sq, scale = sigma.sq)
```

## Simulate our deaths for our two cohorts (y1 and y2)

Here we use Gamma-Gompertz with a common $\rho$ for all ages, but one
could use any baseline and any $\rho(x)$ and the mathematical result
stillholds. An advantage of Gamma-Gompertz is that we can compare our
simulation to the exact analytical result.

```{r}


base.b <- 1/9 # for both cohorts

## for cohort 1

base.a.1 <- 10^-4 

## for cohort 2
## educe base by exp(-rho) of about 10%
rho = .1 # baseline decline 
base.a.2 <- base.a.1 * exp(-rho)

## ages at death
y1 <- rgomp(N,
           b = base.b, ## doesn't vary
           a = base.a.1 * z) ## multiplicative fixed frailty

y2 <- rgomp(N,
           b = base.b, ## doesn't vary
           a = base.a.2 * z) ## multiplicative fixed frailty
```

Now we get the observed life tables for these cohorts using our handy
functions, defined above -- and we calculate the change in hazards we
see in our simulation ("rho.obs.simu").

```{r}
lt.1 <- get.lt(y1)
lt.2 <- get.lt(y2)

## truncate so they have same length
## (they can differ at older ages, depending on chance draws)
n.ages <- max(nrow(lt.1), nrow(lt.2))
lt.1 <- lt.1[1:n.ages,]
lt.2 <- lt.2[1:n.ages,]
x <- 0:(n.ages-1)

## compute observed rho
mx1 <- lt.1$mx
mx2 <- lt.2$mx
rho.obs.simu <- log(mx1/mx2)

plot(x, rho.obs.simu, ylim = c(-.1, .2), type = "o")
abline(h = rho, col = "red")
title("Rates of mortality improvement by age")
```

We see that there is a lot of noise when there are few deaths (early
and late in life). But the main message is that mortality improvement
in our simulation slows at older ages. At age "70" the progress is
only about half of the change in baseline hazards.


## now get predicted rho

Finally we compare to the exact result VM's eq (39*).

```{r}

a1 = base.a.1
a2 = base.a.2
b = base.b
## Gamma-gompertz survival curves
S.bar.1 <- (1 + sigma.sq * (a1/b) * (exp(b*x) - 1))^(1/sigma.sq)
S.bar.2 <- (1 + sigma.sq * (a2/b) * (exp(b*x) - 1))^(1/sigma.sq)

## compute change in S.bar
d.log.S.bar <- log(S.bar.2/S.bar.1) # proportional change in log.S.bar
accent.S.bar <- -d.log.S.bar # minus sign by definition

## compute 
## VM (39*)
## rho.obs = rho.true - sigma.sq * d.log.S.bar
rho.true <- rho
rho.obs.exact =  rho - sigma.sq * accent.S.bar


plot(x, rho.obs.simu, ylim = c(-.1, .2), type = "o")
lines(x, rho.obs.exact, col = "blue")
abline(h = rho, col = "red")
legend("bottom", legend = c("pop rho simulated",
                            "pop rho exact",
                            "baseline rho"),
       pch = c(1, -1,-1),
       lty = c(1,1,1),
       col = c("black", "blue", "red"),
       bty = "n")
abline(h = 0, col = "grey")
```

## Discussion

> Why is simulation noisy at early and late ages?

> Is this like Lee-Carter pattern?

> What is implication for forecasting?

> What is implication for inequality over time?






