---
title: "Reconstruction Ancient European Population Sizes using Batini's sample of Mitochondrial DNA"
output: html_notebook
author: Joshua R. Goldstein
---

In this lab, we use real sequences of mitochondrial DNA to estimate
ancient population sizes. The sequences were made available by Batini
et al, who analyzed sub-populations in their paper using software for
Bayesian inference for comparing groups of individuals. Our approach
here is to use a simpler, but less powerful approach. We will simply
look at pairwise differences of random pairs of individuals. For this
we are not going to do subgroups such as the Greeks or Irish, but are
going to use the entire European sample. The population sizes we
estimate are for the entire population represented by the 328
individuals. 

(Special thanks to Ken Wachter, who taught me this approach and whose
R-code forms the basis for what we will do today.)

# Summary of Batini et al.

Before we begin our own analysis, let's look at the inputs to
Batini's analysis -- the mtDNA haplotype sequences, and then at the
resulting population estimates.

## The mitochondrial DNA


```{r}
source("read_mtdna.R")
haps <- seqs
```

Let's inspect just a bit of one of these sequences
```{r}
print(nchar(haps[213]))
a_segment = substr(haps[213], 1, 100)
print(a_segment)
## tip: don't try to print the 16,000 character whole string. it will clog up your computer.
```


Let's put the haps in a matrix
```{r}
my.list <- strsplit(haps, "")
H <- do.call(cbind, my.list)
print(H[1:10, 1:4]) ## all the same
```

Let's find a site where there's polymorphism
```{r}
hap1 = H[,1]
hap2 = H[,2]
s <- min(which(hap1 != hap2))
head(H[s + -2:2, 1:4]) ## the polymorphic site in context
```

Let's see if that's hap1 is the only "A"
```{r}
table(H[s,])
```

We see that there are 19 individuals with this "A" instead of "G".

> Q: How many pairwise differences in total are there between hap1 and
> hap2?

These are the kind of comparisons _we_ will be doing.

## Ancient population estimates

Let's look at Figure 2 on page 5. 

> Q. What was effective population size in Ireland 1 thousand years ago
> (according to mtDNA)?

> Q. What was effective population size in Ireland 50 thousand years ago
> (according to mtDNA)?


> Q. For the "Irish", what is the annual population growth rate? What's the NRR?

The order of magnitude for each these populations appears to be about 10^4
in  last few KYA and 10^3 50 KYA. Together, perhaps the size is 10
fold. So we're looking at European effective population sizes on the
order of 100,000 the last few thousand years and on the order of
10,000 tens of thousands of years ago.


# Using the Coalescent to estimate changing population size

Our procedure will involve a four steps:

1. Pick 100 _pairs_ of people at random and count their pairwise
   differences.
   
2. Estimate 100 different times of MRCA ($T_i$)  using assumed mutation rate.

3. Estimate $h(t)$, the time-varying hazard of coalescence 

4. Estimate the population size as $Ne(t) = 1/h(t)$.

Let's go!

## 1. Pick 100 _pairs_ of people at random and count their pairwise differences.

```{r}

set.seed(1)
hap_ids = 1:ncol(H)
hap_id_sample = sample(hap_ids,
                       size = 200,
                       replace = FALSE)

hap_id.mat <- matrix(hap_id_sample, 100, 2)

pairwise_diff_fun <- function(hap1, hap2)
{
    h1 <- hap1
    h2 <- hap2
##    h1 <- unlist(strsplit(hap1, ""))
##    h2 <- unlist(strsplit(hap2, ""))
    h1[h1 == "N"] <- NA ## note "N" means missing
    h2[h2 == "N"] <- NA ## making these NA avoids counting as polymorphism
    k = sum(h1 != h2, na.rm = T)
    n_valid = sum(!is.na(h1) & !is.na(h2))
    return(list(k = k, n_valid = n_valid))
}

pairwise_diff_fun(H[,1], H[,2])
```

Now we're ready to do pairwise comparisons of all 100 pairs of
haplotypes.

We'll define the fraction of locii that have mutated (the pairwise differences) as
$$
\bar{Y} = P/C
$$ 

```{r}

P.vec = NULL
C.vec = NULL
for (i in 1:nrow(hap_id.mat))
{
    hap_id.1 = hap_id.mat[i,1]
    hap_id.2 = hap_id.mat[i,2]
    hap1 = H[,hap_id.1]
    hap2 = H[,hap_id.2]
    out = pairwise_diff_fun(hap1, hap2)
    P.vec[i] = out$k
    C.vec[i] = out$n_valid
}

Y.bar = P.vec/C.vec
head(P.vec)
head(C.vec)
head(Y.bar)
```

## 2. Estimate 100 different times of MRCA ($T$)  using assumed mutation rate.

And now we'll use the mutation rate $\theta_m$ given by Batini to
compute time back to MRCA.

First, some background:

Let $a$ index sites and write $Y_a = 1$ if the letters are different,
and $Y_a = 0 otherwise. For 1 site, the probability that it there has been a mutation, is 1
minus the chance that there has been no mutation. 
$$
     P( Y_a = 1 ) = 1 - e^{-T\theta}
     
$$
The mean $\bar{Y}$ across the segment is the proportion of $Y_a$ that
equal 1. So, in expectation for the fraction of sites that mutate is
the same as the probability that 1 site mutates (assuming 
independence of mutation probabilities by site). This allows us to write
$$
     E{\bar{Y}} = 1 - e^{-2T\theta},
$$
where we've added a "2" in order to account that either one of the pairwise branched could have had a mutation, so we our "exposure" is twice the time to MRCA.

Rearranging we get an estimate $\hat{T} of $T$ to be
$$
\hat{T} = {-\log (1 - \bar{Y} ) \over \theta}
$$


Now we're ready to estimate the MRCAs
```{r}
theta_m = 2.21 * 10^(-8) ## Batini page 6 (TMRCA estimation)
T.vec <- -(1/2) * (1/theta_m) * log(1 - Y.bar) ## TMRCAs in years ago
head(T.vec, n = 10)
```

Let's visualize these
```{r}
hist(T.vec)
```

We see a lot of coalescence about 50 KYA, which is as far back as
Batini's estimates go. This means the population was small back
then. 

## 3. Estimate $h(t)$, the time-varying hazard of coalescence 

We'll do this in two ways. First we'll compute the slope of the
logarithm of the survival curve, but we'll see that it is noisy and
needs to be smoothed. Second, we'll construct a "life table" of
coalescence with discrete periods of time.

Let's start with the more continuous version of estimating slopes.

```{r}
## Plot survival curve by order of T
St = (100:1)/100 ## or more generally (length(T.vec):1)/length(T.vec)
t = kya = sort(T.vec)/1000
plot(kya, St, type = "l",
     xlab = "Kilo years ago", ylab = "Fraction of pairs without common ancestor",
     main = "Estimated probability of not coalescing")
plot(kya, log(St), type = "l",
     xlab = "Kilo years ago", ylab = "Log fraction of pairs without common ancestor",
     main = "Etimatated probability of not coalescing, log scale")
     
```

> Q: What is happening to slope in first from 1,000 to 50,000 years
> ago? 

> Q: What does this imply about hazard of coalescence?

> Q: What does this imply about effective population size?


### a) Estimating hazards from smoothed survival curve

First we smooth. 

```{r}
out = lowess(x = kya, y = St, f = 1/5)

St_smooth = out$y
kya_smooth = out$x
plot(kya, St, cex = .5)
lines(kya_smooth, St_smooth, type = 'l')
```

> Q. Is f=1/5 a decent fit? Try a different value.  

Hazards as minus the slope of log

```{r}
haz_hat = -diff(St_smooth)/diff(kya_smooth)
plot(kya_smooth[-1], haz_hat, type = 'l')
```

> Q. Does this plot tell us anything about uncertainty?


### b) Life table appraoch

Now let's estimate the hazards using a "life table", where again
"death" is coalescence and "survival" is still not having a common
ancestor.

```{r}
## we choose these time boundaries arbitrarily ... not sure if
## we'll be able to see the "expansion" after ice age ...
## x = c(0, 12,  20, 40, 65,  180 )*1000    # time interval boundaries
x = c(0,2, 5, 10,  20, 30, 40, 65, 180) * 1000 ## time interval boundaries
```

Define a function to count "exposure" by those pairs that
have MRCA in time intervals
```{r}
get_nax <- function(Ti, x)
{
    ## get person years lived in interval by those who die
    nax <- NULL
    for (i in 1:(length(x)-1))
    {
        s <- Ti >= x[i] & Ti < x[i+1]
        nax[i] = mean(Ti[s] - x[i])
    }
    return(nax)
}
```

Construct the life table

```{r}
n <- diff(x)
T.vec.by.cat <- cut(T.vec, x, include.lowest = T, right = F)
ndx = table(T.vec.by.cat)
lx = rev(cumsum(rev(ndx)))
lxpn = c(lx[-1], 0)
nax = get_nax(Ti = T.vec, x = x)
nLx = n*lxpn + nax * ndx ## exposure
nmx = ndx/nLx ## hazard
lt <- cbind(x = x[-length(x)], n, ndx, lx, nax, nLx, nmx)
print(lt)
```

Let's compare the two estimates

```{r}
x.mid = x[-length(x)] + n/2
plot(x.mid, nmx, type = 'o')
axis(2)
lines(kya_smooth[-1] * 1000, haz_hat/1000, type = "l")
```

We're getting basically the same thing, with a little more hint of
rising hazards (shrinking pop size) in first 20 kya. Nearly the same
thing estimate 40 kya ago, and very little signal before that. 


## 4. Estimate the population size  $N_e(t)$

In order to estimate the population size, we have to think for a minute
about units.

Our hazards are per year, but our logic for why hazards are related to
population size is per generation. Remember, the chance of coalescence
_per generation_ was $1/N$. This means we will want to multiply the annual 
hazard by generation length in order to get hazards per generation unit of time.

Second, mtDNA is inherited only through mothers. So the coalescent
that we are thinking of is only for women. And the effective
population size we're estimating is for females only. We can get a rough estimate of 
both sexes by doubling the number of females.

Putting these two considerations together, we have
$$
\hat{N_e(both sexes)} = 2 \times {1 \over 25 \cdot h(t)},
$$
where the 2 inflates to both sexes, and the 25 inflates the annual
hazard into geneations of 25 years in length.

```{r}

Ne_smooth = 2 / (haz_hat/1000 * 25)

Ne_lifetable = 2 / (nmx * 25)

## create step function for plotting
Ne_lifetable_step = rep(Ne_lifetable, n/1000)
kya_step = 1:(max(x)/1000)

```

Plotting the results

```{r}
plot(kya_smooth[-1], Ne_smooth, type = 'l', ylim = c(1000, 60000), log = 'y',
     lty = 2)
lines(kya_step, Ne_lifetable_step, type = 'l', lwd = 2)
```

## [1] 31732.155  8581.928  4999.394  1299.908  1968.916

## Uncertainty

> Q. How could we evaluate undercertainty? (Hint: resampling)

> Q. What does this estimate leave out?

(We probably won't do this in class. If we don't give it a try at
home.)



Let's loop through and do the whole estimation 40 times

```{r}
n_trials = 40
Ne.mat <- matrix(NA, nrow = n_trials, ncol = max(x)/1000)
set.seed(1)
for (r in 1:n_trials)
{
    print(r)
    ## sample
    hap_id_sample = sample(hap_ids,
                           size = 200,
                           replace = FALSE)
    hap_id.mat <- matrix(hap_id_sample, 100, 2)
    ## estimate MRCA distribution
    P.vec = NULL
    C.vec = NULL
    for (i in 1:nrow(hap_id.mat))
    {
        hap_id.1 = hap_id.mat[i,1]
        hap_id.2 = hap_id.mat[i,2]
        hap1 = H[,hap_id.1]
        hap2 = H[,hap_id.2]
        out = pairwise_diff_fun(hap1, hap2)
        P.vec[i] = out$k
        C.vec[i] = out$n_valid
    }
    Y.bar = P.vec/C.vec
    T.vec <- -(1/2) * (1/theta_m) * log(1 - Y.bar) ## TMRCAs in years ago
    ## estimate Ne
    T.vec.by.cat <- cut(T.vec, x, include.lowest = T, right = F)
    ndx = table(T.vec.by.cat)
    lx = rev(cumsum(rev(ndx)))
    lxpn = c(lx[-1], 0)
    nax = get_nax(Ti = T.vec, x = x)
    nLx = n*lxpn + nax * ndx ## exposure
    nmx = ndx/nLx ## hazard
    Ne_lifetable = 2 / (nmx * 25)
    ## create step function for plotting
    Ne_lifetable_step = rep(Ne_lifetable, n/1000)
    kya_step = 1:(max(x)/1000)
    ## save result
    Ne.mat[r,] <- Ne_lifetable_step
}

Ne.interval <- apply(Ne.mat, 2, quantile, c(.1,.5, .9))
matplot(t(Ne.interval), type = 'l', log = 'y', col = "grey", lty = 1, lwd = 2)
lines(Ne.interval["50%",], lwd = 4)
```

So it seems fairly clear that effective population size has been
growing the last 50 thousand years, from a low of a few thousand to a
few tens of thousand. 

The general trend in growth is consistent with Batini but

* Our total population size seems smaller by a factor of about 4 or 5.

* We don't have the resolution to see increase between 10 and 20 kya,
  the end of the Last Glacial Maximum or the more recent Bronze Age
  steppe expansion 2 to 5 kya.
  
In order to get more resolution and study sub-group differences, we
would want to turn to methods that do more than pair-wise comparisons,
giving us more detailed information about the effective population
sizes of the past.


## Some exercises

1. Try doing 150 pairwise comparisons (instead of 100) and see what
   happens.

2. Try doing 50 pairwise comparisons (instead of 100) and see what
   happens.
   
3. Try changing the interval sizes (or the smoothing parameter) to see
   if you can see either the post-glaciation population increase or
   the Bronze Age increase? (I don't know if it's possible with our
   methods.)
   
4. *possible term project* Figure out how to use BEAST or some other software and
   reproduce their results for one sub-population and tell us what you learned.
   
   

   


